---
title: "Sample Replicate Correlation Studies for Example Experiment"
date: "Jan. 27, 2018"
authhor: "C.M. Monaco"
output:
  html_notebook:
    code_folding: hide
---

This document contains post-normalization quality control steps for identification of low correlation bio-replicates using pairwise Pearson Correlation tests.

# How to Use This Script

This script is meant to be run on normalized count data prior to averaging to identify any possible bio-replicates that should be discarded. Currently this process is performed manually and each bio-replicate group must be coded for comparison. Once a standard naming convention for samples is established, this process can be automated. The script should be edited where indicated by an `## EDIT ##` and creating more groups can be performed by simply copying and pasting the plotting function calls.

Outlying samples are indicated with red p-values on the generated plots.

```{r}
## EDIT ##
# Begin by reading in normalized counts from a CSV file.
counts <- data.frame(read.table("norm_counts_output/norm_counts.csv", # <- change path to location of norm
                                header = TRUE,
                                sep = ",",
                                row.names = 1))
```

# Biological Replicate Pairwise Correlation Study

Here, sets of bio-reps are pairwise correlated in order to find replicates that do not correlate well. Poorly correlated replicates should not be averaged into the final data set.

```{r}
## EDIT ##
# Replicate Groupings
# --------------------
# Typically sample names in normalized counts file will be in
# alphabetical order; meaning that some sort of grouping can be established
# that can aid in quickly creating correlation plots. 
# For example:
# SA_1 | SA_2 | SA_3 | SA_4 | SB_1 | SB_2 ...
# So here to capture group A we would use 1:4 and to capture group B we would use (1:4)*2
# In this example there are different groups with sub groupings
gr1 <- 1:4
gr2 <- 5:7
gr3 <- 8:10

## DO NOT EDIT THIS FUNCTION ##
# performs the Pearson correlation test and prints results to plots.
panel.cor <- function(x, y, digits = 2, cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  # correlation coefficient
  r <- cor(x, y, method = "pearson")
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste("r= ", txt, sep = "")
  text(0.5, 0.6, txt)
  # p-value calculation
  p <- cor.test(x, y)$p.value
  txt2 <- format(c(p, 0.123456789), digits = digits)[1]
  txt2 <- paste("p= ", txt2, sep = "")
  if(p<0.01)
  {
    txt2 <- paste("p= ", "<0.01", sep = "")
    text(0.5, 0.4, txt2)
  }
  else
  {
    text(0.5, 0.4, txt2, col = "red")
  }
}
```

## Ld Group 1

```{r}
## EDIT ##
# This function creates the correlation plots. This section of code should be 
# copied and pasted for each group.
i <- 0
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Ld Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Ld Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Le Group 1

```{r}
i <- 1
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Le Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Le Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Lm Group 1

```{r}
i <- 2
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Lm Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Lm Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Sd Group 1

```{r}
i <- 3
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Sd Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Sd Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Se Group 1

```{r}
i <- 4
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Se Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Se Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Sm Group 1

```{r}
i <- 5
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Sm Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Sm Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Vd Group 1

```{r}
i <- 6
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Vd Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Vd Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Ve Group 1

```{r}
i <- 7
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Ve Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Ve Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```

## Vh Group 1

```{r}
i <- 8
pairs(counts[,gr1+10*i], upper.panel = panel.cor)
```

## Vh Group 2

```{r}
pairs(counts[,gr2+10*i], upper.panel = panel.cor)
```

## Vh Group 3

```{r}
pairs(counts[,gr3+10*i], upper.panel = panel.cor)
```