---
title: "Run-Away Barcode Identification"
date: "Jan. 27, 2018"
authhor: "C.M. Monaco"
output:
  html_notebook:
    code_folding: hide
---

This document contains post-normalization quality control steps for identification of possible run-away barcode sequences.

# How to Use This Script

This script is meant to be run on normalized count data prior to averaging to identify any possible run-away barcodes. The script should be edited where indicated by an `## EDIT ##`.

```{r}
## EDIT ##
# Begin by reading in normalized counts from a CSV file.
counts <- data.frame(read.table("counts/norm_counts.csv", # <- change path to location of counts file
                                header = TRUE,
                                sep = ",",
                                row.names = 1))
```

# Multigroup Barcode Outlier Identification

## Single Linkage Hierarchical Clustering

Outliers will appear as outgroups in the clustering dendrogram.

```{r, fig.width=10}
d <- dist(counts, method = "euclidean")
h <- hclust(d, method = "single")
plot(h, hang = -1, cex=0.5)
```

## Principle Component Analysis

Outliers will appear as clusters seperated from the main clustering of Barcodes.

```{r}
counts.pca <- prcomp(counts, center = TRUE)

library(ggfortify)
autoplot(counts.pca, label = TRUE, label.size = 3)
```

## Outlier Heatmaps (No Clustering)

This uncluseterd heatmap will show evidence of a run-way by showing any barcodes with disproportionaly hight number of reads across nearly all samples.

```{r, fig.width=10}
library(ComplexHeatmap)
library(circlize)
Heatmap(counts,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        col = colorRamp2(c(0,2.5,5), c("red","white","blue")),
        column_names_gp = gpar(cex = 0.5),
        row_names_gp = gpar(cex = 0.5)
        )
```